(note: you must first run "functions.r" then "Logistic_Regressions_testing_Conti_GRS.R")

## Part 1: Asymptomatic Screening ##

Can the GRS predict which men in the UKB will be diagnosed with prostate cancer in a given timeframe after the assessment centre?

# Exclusion Criteria #

*General:*

Females (based on self-report of “Sex”. Using “Genetic Sex” did not alter the results)

Anyone without linked genetic data (imputed genotypes)

Anyone with a record of prostate cancer diagnosis before initial assessment centre visit

*Controls:*

Anyone with any of the following ICD9 codes in their Hospital Episode Statistics (HES) records: 

185: Malignant neoplasm of prostate
V10.4: Personal history of malignant neoplasm of genital organs
233.4: Carcinoma in situ of the prostate 
236.5: Neoplasm of uncertain behavior of the prostate 
602.3: Dysplasia of the prostate 
602.1: Transurethral (ultrasound) guided laser induced prostatectomy (TULIP)
602.9: Other transurethral prostatectomy 
603: Suprapubic prostatectomy
604: Retropubic prostatectomy
605: Radical prostatectomy
606.1: Local excision of lesion of prostate 
606.2: Perineal prostatectomy
606.9: Other prostatectomy

Anyone with any of the following ICD10 codes in their HES records or in the cancer registry:

C61: Malignant neoplasm of prostate
Z85.4: Personal History of malignant neoplasm of genital organs
R97.2: Elevated prostate specific antigen [PSA] 
D07.5: Carcinoma in situ of prostate
D40.0: Neoplasm of uncertain behavior of prostate
N42.3: Dysplasia of prostate

Anyone with any of the following codes in their OPCS records: 

M61: Prostatectomy
M61.1: Radical prostatectomy
M61.2: Retropubic Prostatectomy
M61.3: Transvesical Prostatectomy
M61.4: Perineal Prostatectomy
X65: Radiotherapy Delivery
X67: Preparation of radiotherapy
X68: Brachytherapy preparation
M70.6: Radioactive seed implantation into prostate
Y35: Introduction Material Radioactive Removable NOC
Y36: Introduction Material Non-removable NOC
T85.6: Block dissection of pelvic lymph nodes
M70.2: Perineal needle biopsy of prostate
M70.3: Rectal needle biopsy of prostate
N04: Orchidectomy
M65: Endoscopic resection of prostate
M68: Endoscopic insertion of prosthesis into prostate
M67.1: Endoscopic cryotherapy to lesion of prostate
M71.1: High intensity focused ultrasound of prostate
M71.2: Implantation of radioactive substance into prostate

# Conditions #

Age = Age at initial assessment centre visit

Genetic similarity* is defined by Pan-UKB study (Karczewski et al., 2025)

4 SNPs missing from GRS calculation, including rs72725854

Prostate cancer cases defined as:
- At least one “C61” or “185” event in Hospital Records (only if the diagnosis is confirmed in the cancer registry or death records)
- At least one “C61” event in Cancer Registry
- At least one “C61” event in death records as “Cause of Death”

Ethnicity is based on self-report and clustered as follows:

- White = White, British, Irish, Any other White background
- Black = African, Caribbean, Any other Black background, Black or Black British
- Mixed = White and Black African, White and Black Caribbean
- Black + Mixed = African, Caribbean, Any other Black background, Black or Black British, White and Black African, White and Black Caribbean
- South Asian = Pakistani, Indian, Bangladeshi
- East Asian = Chinese
- Other = Other ethnic group, Any other mixed background, Prefer not to answer, Do not know, Mixed, NA, Any other Asian background, White and Asian, Asian or Asian British

Note: Genetic similarity is the preferred term for what is called “Genomic Ancestry” in the UK Biobank

# Logistic Regression Table 1: How do Conti's Multi-ancestry and African-specific GRSs perform in predicting prostate cancer diagnosis any time after the assessment centre? #

```{r, echo=FALSE}

table1 <- formatted %>%                        
  dplyr::filter(                          
    Population == "White" | Population == "Black" | Population == "Mixed" | Population == "Black + Mixed",
    Predictor == "ContimultiethnicGRS267" | Predictor == "ContiAfricanGRS246",
    Outcome == "PrCa",
    Covariates == "None"
  )    

print(table1)

```

# Logistic Regression Table 2: How do Conti's Multi-ancestry and African-specific GRSs perform in predicting prostate cancer diagnosis any time after the assessment centre when combined with Age? #

```{r, echo=FALSE}

table2 <- formatted %>%                        
  dplyr::filter(                          
    Population == "White" | Population == "Black" | Population == "Mixed" | Population == "Black + Mixed",
    Predictor == "ContimultiethnicGRS267" | Predictor == "ContiAfricanGRS246",
    Outcome == "PrCa_10yrs",
    Covariates == "Age"
  )

print(table2)

```

Conclusion: Age always helps, and using the African-specific weightings makes very little difference



# Logistic Regression Table 3: How do Conti's Multi-ancestry and African-specific GRSs perform in predicting prostate cancer diagnosis within 10 years after the assessment centre when combined with Age? #

```{r, echo=FALSE}

table3 <- formatted %>%                        
  dplyr::filter(                          
    Population == "White" | Population == "Black" | Population == "Mixed" | Population == "Black + Mixed" | Population == "AFR",
    Predictor == "ContimultiethnicGRS267",
    Outcome == "PrCa_10yrs",
    Covariates == "Age"
  )

print(table3)

```

But what does "Genetic Similarity" even mean?

## Genetic Similarity ##

- Defined in Pan-UKB by Karczewski et al. (2025)

- Statistically compared Biobank participants to HGDP + 1000 genomes

- Is now the Biobank’s built-in “Genomic Ancestry” variable

- Problem: strict clustering method that excluded >50,000 participants
   - Anyone who did not have >50% probability of fitting into a particular cluster (EUR, AFR, EAS, CSA, MID, AMR) was excluded


## Principal Components plot 1: HGDP+1KG Reference Panel ##

```{r, echo=FALSE}

library(ggplot2)

dxdownload("Callum/HGDP_1KG/HGDP_1KG_PCs.csv") # HGDP+1KG Principal components as calculated using AncestryProbability1_bigsnpr folder scripts
dxdownload("Callum/HGDP_1KG/release_3.1_secondary_analyses_hgdp_1kg_v2_metadata_and_qc_gnomad_meta_updated.tsv") # This can be downloaded from https://console.cloud.google.com/storage/browser/gcp-public-data--gnomad/release/3.1/secondary_analyses/hgdp_1kg_v2/metadata_and_qc
dxdownload("Callum/HGDP_1KG/hgdp_tgp_pca_covid19hgi_snps_scores.txt") # can be downloaded from:
# https://console.cloud.google.com/storage/browser/covid19-hg-public/pca_projection;tab=objects?prefix=&forceOnObjectsSortingFiltering=false

## Working out which column of the HGDP+1KG metadata file to use as ancestry labels

HGDP_1KG_metadata <- read.delim("release_3.1_secondary_analyses_hgdp_1kg_v2_metadata_and_qc_gnomad_meta_updated.tsv")

## Decided on using "Genetic Region" as it is closest to what we see in UK Biobank "Genomic Ancestry" variable

HGDP_1KG_metadata_ancestrylabel <- HGDP_1KG_metadata %>%
  dplyr::select(c("s", "hgdp_tgp_meta.Genetic.region")) %>%
  dplyr::rename("IID" = "s", "Genomic_ancestry" = "hgdp_tgp_meta.Genetic.region")

HGDP_1KG_PCs <- read.csv("HGDP_1KG_PCs.csv") # PCs as I calculated them
HGDP_1KG_PCs_w_labels <- merge(HGDP_1KG_PCs, HGDP_1KG_metadata_ancestrylabel, by = "IID")

HGDP_1KG_PCs_og <- read.delim("hgdp_tgp_pca_covid19hgi_snps_scores.txt") %>% # Publicly available PCs
  dplyr::rename("IID" = "s")
HGDP_1KG_PCs_og_w_labels <- merge(HGDP_1KG_PCs_og, HGDP_1KG_metadata_ancestrylabel, by = "IID")

################################################################################
#-------------------------------- Setup done ----------------------------------#
################################################################################

test_population <- HGDP_1KG_PCs_og_w_labels              # Publicly available PCs
#test_population <- HGDP_1KG_PCs_w_labels                 # PCs as I calculated them
#test_population <- PCa_iv_covariates_GRS_clean           # Projected PCs of UKB participants - requires running "Logistic_Regresions_testing_Conti_GRS.R" to create

HGDP_1KG_PC_plot <- ggplot(test_population, aes(x = PC1, y = PC2, color = Genomic_ancestry)) +
  geom_point(size = 0.5) +
  theme_minimal() +
  # coord_cartesian(xlim = c(-50, 80), ylim = c(-50, 60)) +
  labs(
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Genomic ancestry"
  )

print(HGDP_1KG_PC_plot)

```

This is the reference panel (HGDP+1KG)

## Principal Components plot 2: UK Biobank Participants who were successfully clustered ##

```{r, echo=FALSE}

library(ggplot2)

dxdownload("Callum/HGDP_1KG/HGDP_1KG_PCs.csv") # HGDP+1KG Principal components as calculated using AncestryProbability1_bigsnpr folder scripts
dxdownload("Callum/HGDP_1KG/release_3.1_secondary_analyses_hgdp_1kg_v2_metadata_and_qc_gnomad_meta_updated.tsv") # This can be downloaded from https://console.cloud.google.com/storage/browser/gcp-public-data--gnomad/release/3.1/secondary_analyses/hgdp_1kg_v2/metadata_and_qc
dxdownload("Callum/HGDP_1KG/hgdp_tgp_pca_covid19hgi_snps_scores.txt") # can be downloaded from:
# https://console.cloud.google.com/storage/browser/covid19-hg-public/pca_projection;tab=objects?prefix=&forceOnObjectsSortingFiltering=false

## Working out which column of the HGDP+1KG metadata file to use as ancestry labels

HGDP_1KG_metadata <- read.delim("release_3.1_secondary_analyses_hgdp_1kg_v2_metadata_and_qc_gnomad_meta_updated.tsv")

## Decided on using "Genetic Region" as it is closest to what we see in UK Biobank "Genomic Ancestry" variable

HGDP_1KG_metadata_ancestrylabel <- HGDP_1KG_metadata %>%
  dplyr::select(c("s", "hgdp_tgp_meta.Genetic.region")) %>%
  dplyr::rename("IID" = "s", "Genomic_ancestry" = "hgdp_tgp_meta.Genetic.region")

HGDP_1KG_PCs <- read.csv("HGDP_1KG_PCs.csv") # PCs as I calculated them
HGDP_1KG_PCs_w_labels <- merge(HGDP_1KG_PCs, HGDP_1KG_metadata_ancestrylabel, by = "IID")

HGDP_1KG_PCs_og <- read.delim("hgdp_tgp_pca_covid19hgi_snps_scores.txt") %>% # Publicly available PCs
  dplyr::rename("IID" = "s")
HGDP_1KG_PCs_og_w_labels <- merge(HGDP_1KG_PCs_og, HGDP_1KG_metadata_ancestrylabel, by = "IID")

################################################################################
#-------------------------------- Setup done ----------------------------------#
################################################################################

#test_population <- HGDP_1KG_PCs_og_w_labels              # Publicly available PCs
#test_population <- HGDP_1KG_PCs_w_labels                 # PCs as I calculated them
test_population <- PCa_iv_covariates_GRS_clean           # Projected PCs of UKB participants - requires running "Logistic_Regresions_testing_Conti_GRS.R" to create

UKB_PC_plot1 <- ggplot(test_population %>% dplyr::filter(!is.na(Genomic_ancestry)), aes(x = PC1, y = PC2, color = Genomic_ancestry)) +
  geom_point(size = 0.5) +
  theme_minimal() +
  # coord_cartesian(xlim = c(-50, 80), ylim = c(-50, 60)) +
  labs(
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Genomic ancestry"
  )

print(UKB_PC_plot1)

```

## Principal Components plot 3: all UK Biobank Participants ##

```{r, echo=FALSE}

library(ggplot2)

dxdownload("Callum/HGDP_1KG/HGDP_1KG_PCs.csv") # HGDP+1KG Principal components as calculated using AncestryProbability1_bigsnpr folder scripts
dxdownload("Callum/HGDP_1KG/release_3.1_secondary_analyses_hgdp_1kg_v2_metadata_and_qc_gnomad_meta_updated.tsv") # This can be downloaded from https://console.cloud.google.com/storage/browser/gcp-public-data--gnomad/release/3.1/secondary_analyses/hgdp_1kg_v2/metadata_and_qc
dxdownload("Callum/HGDP_1KG/hgdp_tgp_pca_covid19hgi_snps_scores.txt") # can be downloaded from:
# https://console.cloud.google.com/storage/browser/covid19-hg-public/pca_projection;tab=objects?prefix=&forceOnObjectsSortingFiltering=false

## Working out which column of the HGDP+1KG metadata file to use as ancestry labels

HGDP_1KG_metadata <- read.delim("release_3.1_secondary_analyses_hgdp_1kg_v2_metadata_and_qc_gnomad_meta_updated.tsv")

## Decided on using "Genetic Region" as it is closest to what we see in UK Biobank "Genomic Ancestry" variable

HGDP_1KG_metadata_ancestrylabel <- HGDP_1KG_metadata %>%
  dplyr::select(c("s", "hgdp_tgp_meta.Genetic.region")) %>%
  dplyr::rename("IID" = "s", "Genomic_ancestry" = "hgdp_tgp_meta.Genetic.region")

HGDP_1KG_PCs <- read.csv("HGDP_1KG_PCs.csv") # PCs as I calculated them
HGDP_1KG_PCs_w_labels <- merge(HGDP_1KG_PCs, HGDP_1KG_metadata_ancestrylabel, by = "IID")

HGDP_1KG_PCs_og <- read.delim("hgdp_tgp_pca_covid19hgi_snps_scores.txt") %>% # Publicly available PCs
  dplyr::rename("IID" = "s")
HGDP_1KG_PCs_og_w_labels <- merge(HGDP_1KG_PCs_og, HGDP_1KG_metadata_ancestrylabel, by = "IID")

################################################################################
#-------------------------------- Setup done ----------------------------------#
################################################################################

#test_population <- HGDP_1KG_PCs_og_w_labels              # Publicly available PCs
#test_population <- HGDP_1KG_PCs_w_labels                 # PCs as I calculated them
test_population <- PCa_iv_covariates_GRS_clean           # Projected PCs of UKB participants - requires running "Logistic_Regresions_testing_Conti_GRS.R" to create

UKB_PC_plot1 <- ggplot(test_population, aes(x = PC1, y = PC2, color = Genomic_ancestry)) +
  geom_point(size = 0.5) +
  theme_minimal() +
  # coord_cartesian(xlim = c(-50, 80), ylim = c(-50, 60)) +
  labs(
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Genomic ancestry"
  )

print(UKB_PC_plot1)

```

This is what it looks like when you include participants who were not successfully clustered.

## Principal Components plot 4: UK Biobank Participants by self-reported ethnicity ##

```{r, echo=FALSE}

library(ggplot2)

dxdownload("Callum/HGDP_1KG/HGDP_1KG_PCs.csv") # HGDP+1KG Principal components as calculated using AncestryProbability1_bigsnpr folder scripts
dxdownload("Callum/HGDP_1KG/release_3.1_secondary_analyses_hgdp_1kg_v2_metadata_and_qc_gnomad_meta_updated.tsv") # This can be downloaded from https://console.cloud.google.com/storage/browser/gcp-public-data--gnomad/release/3.1/secondary_analyses/hgdp_1kg_v2/metadata_and_qc
dxdownload("Callum/HGDP_1KG/hgdp_tgp_pca_covid19hgi_snps_scores.txt") # can be downloaded from:
# https://console.cloud.google.com/storage/browser/covid19-hg-public/pca_projection;tab=objects?prefix=&forceOnObjectsSortingFiltering=false

## Working out which column of the HGDP+1KG metadata file to use as ancestry labels

HGDP_1KG_metadata <- read.delim("release_3.1_secondary_analyses_hgdp_1kg_v2_metadata_and_qc_gnomad_meta_updated.tsv")

## Decided on using "Genetic Region" as it is closest to what we see in UK Biobank "Genomic Ancestry" variable

HGDP_1KG_metadata_ancestrylabel <- HGDP_1KG_metadata %>%
  dplyr::select(c("s", "hgdp_tgp_meta.Genetic.region")) %>%
  dplyr::rename("IID" = "s", "Genomic_ancestry" = "hgdp_tgp_meta.Genetic.region")

HGDP_1KG_PCs <- read.csv("HGDP_1KG_PCs.csv") # PCs as I calculated them
HGDP_1KG_PCs_w_labels <- merge(HGDP_1KG_PCs, HGDP_1KG_metadata_ancestrylabel, by = "IID")

HGDP_1KG_PCs_og <- read.delim("hgdp_tgp_pca_covid19hgi_snps_scores.txt") %>% # Publicly available PCs
  dplyr::rename("IID" = "s")
HGDP_1KG_PCs_og_w_labels <- merge(HGDP_1KG_PCs_og, HGDP_1KG_metadata_ancestrylabel, by = "IID")

################################################################################
#-------------------------------- Setup done ----------------------------------#
################################################################################

#test_population <- HGDP_1KG_PCs_og_w_labels              # Publicly available PCs
#test_population <- HGDP_1KG_PCs_w_labels                 # PCs as I calculated them
test_population <- PCa_iv_covariates_GRS_clean           # Projected PCs of UKB participants - requires running "Logistic_Regresions_testing_Conti_GRS.R" to create

UKB_PC_plot1 <- ggplot(test_population %>% dplyr::filter(ethnicity_group_narrow == "White" | ethnicity_group_narrow == "Black" | ethnicity_group_narrow == "Mixed"), aes(x = PC1, y = PC2, color = ethnicity_group_narrow)) +
  geom_point(size = 0.5) +
  theme_minimal() +
  # coord_cartesian(xlim = c(-50, 80), ylim = c(-50, 60)) +
  labs(
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Self-reported ethnicity"
  )

print(UKB_PC_plot1)

```


